Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-18 cordillera-cycle.tex
Polish and add missing parts.

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving atmospheric
  forcing, all simulations produce two major glaciation events during
  marine isotope stages~4 (61.9--55.4\,kyr) and~2
  (29.5--16.9\,kyr). The timing of glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional oceanic sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena Mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remains of Cordilleran ice during the
  early Holocene (10.9--9.5\,kyr).

@@

Ice sheet dynamics are typically modelled using a combination of internal
deformation and basal sliding. PISM is a~shallow ice sheet model, which implies
that the balance of stresses is approximated based on their predominant
components. The Shallow Shelf Approximation (SSA) is used as a ``sliding law''
for the Shallow Ice Approximation (SIA) by adding velocity solutions of the
two approximation
\citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}\renote{
    Arjen, I don't agree with your comment. Reference to equation number(s)
    show the reader where to look. \citet{Winkelmann.etal.2011} is a rather
    long paper that also treats many other aspects of the model. Their SIA+SSA
    approach builds on \citep{Bueler.Brown.2009}, but it is slightly different.
    It is the one in use in PISM, and in my simulations.}.
Ice rheology depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
the atmospheric forcing (Sect.~\ref{sec:atm}) provides the upper boundary
condition to the ice enthalpy model. Temperature is computed subglacially to
a~depth of 3\,km\renote{
    Yes, I changed this from 1 to 3\,km as we now model longer period of times
    than in the previous paper. The penetration depth of temperature anomalies
    (i.e. skin depth) depends on the period of temperature oscillations at the
    surface. The longer the period (in our case 100\,kyr), the deeper the
    temperature change. A good way to estimate what this parameter should be is
    to solve the 1-D temperature diffusion equation in the case of periodic
    surface forcing, using rock heat capacity and rock thermal conductivity
    values similar as in the numerical model. This basically forms the basis
    for my choice of the 3\,km value.},
where it is conditioned by a lower boundary geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
high spatial geothermal variability in the region
\citep{Blackwell.Richards.2004}, it is, on average, representative of available
heat flow measurements. In the low-resolution simulations, the vertical grid
consists of up to 51~enthalpy layers in the ice sheet and 31~temperature layers
in the bedrock. In the high-resolution simulations, up to 101~ice layers, and
61~bedrock layers are used.

@@

However, the exact timing of modelled MIS~2 maxima depends strongly on the
choice of applied palaeo-temperature record, which allows for a more in-depth
comparison with geological evidence for the timing of maximum Cordilleran ice
sheet extent. In the Puget lowland (Fig.~\ref{fig:locmap}), the LGM advance of
the southern Cordilleran ice sheet margin has been constrained by radiocarbon
dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka\,BP}
\citep{Porter.Swanson.1998}.
These dates have been confirmed by the offshore sedimentary record, which shows
an increase of glaciomarine sedimentation between 19.5 and
16.2\,\unit{\chem{^{14}C}\,cal\,ka\,BP} \citep{Cosma.etal.2008}. Radiocarbon
dating of the northern Cordilleran ice sheet margin is much less constrained
but straddles presented constraints from the southern margin. However,
cosmogenic exposure dating
places the timing of maximum CIS extent during the McConnell glaciation close
to 17\,ka \citep{Stroeven.etal.2010, Stroeven.etal.2014}.

Among the simulations presented here, only those forced with the GRIP, EPICA
and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
extent that may be compatible with these field constraints
(Fig.~\ref{fig:timeseries}, lower panel; Table~\ref{tab:extrema}).
Simulations driven by the NGRIP, ODP~1012 and ODP~1020
palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
ice sheet extents that pre-date field-based constraints by several thousands of
years. Concerning the simulations driven by oceanic records, this early
deglaciation is caused by an early warming present in the alkenone
palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
\citealp[Fig.~3]{Herbert.etal.2001}). However, this
early warming is a local effect, corresponding to a weakening of the California
current \citep{Herbert.etal.2001}. The California current, driving cold
waters southwards along the south-western coast of North America\renote{
    Not only USA but also Baja California in Mexico.},
has been shown to have weakened during each peak global glaciation (in SPECMAP)
durint the past 550\,ka, including the LGM, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
\citep{Herbert.etal.2001}.

@@

\todo{Write something about NGRIP.}

Hence, we focus the rest of our analysis on simulations forced by
palaeo-temperature records from GRIP and EPICA ice cores. To allow for a more
detailed comparison against the geologic record, these two simulations were
re-run on the high-resolution grid (Fig.~\ref{fig:timeseries}, lower panel,
dotted lines).

@@

The modelled ice sheet configurations corresponding to ice volume maxima during
MIS~4 are more variable (Fig.~\ref{fig:snapshots}, upper panels;
Fig.~\ref{fig:mis4}). The GRIP simulation (Fig.~\ref{fig:mis4}, left panel)
results in a modelled maximum ice sheet that closely resemble that obtained
during MIS~2, with the only major differences being a slightly less developed
northern and eastern sectors. However, the EPICA simulations produce a reduced
ice volume maximum (Fig.~\ref{fig:timeseries}), which translates in the
modelled ice sheet geometry into a much less developed southern sector, more
restricted ice cover on the northern and eastern flanks, and generally lower
ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).

In the GRIP simulation, some sections of the MIS~4 maximum ice margin are
slightly more extensive than their MIS~2 counterparts. This is notably the case
in the Puget Lowland, along parts of the marine margin, and parts of the north
slope of the Alaska Range. However, these differences generally correspond to a
few model grid cells. Their interpretation would be hazardous considering
the simplicity of our atmospheric forcing, and the scale of our simulations.

@@

It is generally believed that the Cordilleran ice sheet formed by the
coalescence of of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
In our simulations, major ice-dispersal centres, visible on the modelled ice
cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
Mountains (CM), the Columbia and Rocky mountains (CR), the Skeena Mountains
(SM), and the Selwyn and MacKenzie Mountains (SMK). Although the Coast,
Columbia, Rocky and Skeena Mountains (CM, CR, SM) are covered by mountain
glaciers for most of the model time, providing durable nucleation centre for
the initiation of ice sheet growth, this is not the case for the Selwyn and
MacKenzie mountains (SMK), where ice cover on the highest peaks is limited to a
fraction of the last glacial cycle. In other words, the Selwyn and MacKenzie
mountains appear as a secondary ice-dispersal centre which ``activates'' only
during the coldest periods of the last glacial cycle. The Northern Rocky
Mountains (Fig.~\ref{fig:duration}, NR), do not act as a nucleation centre,
but rather as a pinning point for the ice margin coming from the west.

Perhaps the most striking feature displayed by the distributions of modelled
ice cover is the persistence of the Skeena Mountains ice cap throughout the
entire last glacial cycle and its predominance over other ice-dispersal centres
(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}). Regardless of the applied
forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:snapshots},
middle panels), and serves as a nucleation centre at the onset of the glacial
readvance towards the LGM (MIS~2). The importance of residual ice for North
American glacial history leading up to the LGM has been illustrated by the
MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}\renote{
    Does this sentence really fit here? I feel that this sudden mention of the
    Laurentide ice sheet perturbates a bit the flow. Moreover, knowing that
    the Laurentide ice sheet is probably the main driver for the 100\,ka
    sea-level cycle, it would have been surprising if it had completely
    vanished at MIS~3. But for the Cordilleran ice sheet, usually thought of
    as fast-responding and intermittent, I feel that the presence of residual
    ice is less trivial.}.

@@

The modelled distribution of warm-based ice cover (Figs.~\ref{fig:warmbase}
and~\ref{fig:warmfrac}) is inevitably affected by our assumption of a constant,
70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
(Sect.~\ref{sec:overview}). However, both the Skeena Mountain area and the area
west of the MacKenzie Mountains are areas of higher-than-average geothermal
heat flux \citep{Blackwell.Richards.2004}. We can therefore expect a higher
frequency of warm-based ice cover in this areas if we were to include
spatially-variable geothermal forcing in our simulations.

@@

In the North American Cordillera, the presence of lateral melt-water channels
at high elevation \citep{Margold.etal.2011, Margold.etal.2013a,
Margold.etal.2014}, and abundant esker systems at low elevation
\citep{Burke.etal.2012, Burke.etal.2012a, Perkins.etal.2013,
Margold.etal.2013}\footnote{
    Is this a good choice of references?}
indicate a rapid deglaciation. The southern and northern margins of the
Cordilleran ice sheet reached their last glacial maximum extent around
17\,ka \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998, Cosma.etal.2008,
Stroeven.etal.2010, Stroeven.etal.2014}, putting an older bound to the
onset of ice retreat. However, he timing of the late deglaciation is less well
constrained, but recent cosmogenic dates from north-central British Columbia
indicate that a seizable ice cap emanating from the central Coast Mountains or
the Skeena Mountains persisted into the Younger Dryas chronozone, at least
until 12.4\,ka \citep{Margold.etal.2014}.

@@

Because a general tenant in glacial geomorphology is that the majority of
landforms (lineations and eskers) are part of the deglacial envelop
\citep[terminology from]{Kleman.etal.2006}, that is that these were formed
close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
Kleman.etal.1997, Kleman.etal.2010}, we present a map of basal flow directions
immediately preceding deglaciation or at the time of cessation of sliding
inside the cold-base retreating margin (Fig.~\ref{fig:lastflow})
\todo{description of the general trends that the map shows}. Patterns of
glacial lineation formed in the northern and southern sectors of the
Cordilleran ice sheet and in the Liard Lowland (\citealp{Prest.etal.1968};
\citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010};
\citealp[Fig.~2]{Margold.etal.2013}), similarities with the patterns of
deglacial ice flow from modelling (Fig.~\ref{fig:lastflow}). However, this is
not the case for the Interior Plateau of British Columbia, where both
simulations predict negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), but where an impressive set of glacial delineations
indicate a substantial eastward flow component of the Cordilleran ice sheet
\citep{Prest.etal.1968, Kleman.etal.2010}.

The Interior Plateau lineation swarm could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that the modelled LGM ice sheet is too thick, or
that the ice divide is positioned too far to the east \todo{descript the
problem for the incongruent results of these two conditions}. Because feedback
mechanisms between ice sheet topography and regional climate are absent in our
model, including wind redirection, orographic precipitation effects, and latent
warming of moisture-depleted air, we regard misfeeds in ice thickness and ice
devide location as plausible \citep{Seguinot.etal.2014}. Another explanation
for the incongruent results could be that the Interior Plateau lineation swarm
predates deglaciation and that deglacitation landforms are largely absent. The
modelled deglaciation of the Interior Plateau, central British Columbia,
consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
southwards-flowing (Fig.~\ref{fig:lastflow}) non-sliding ice lobes positioned
in-between deglaciated (ice-free) mountain ranges
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
appears compatible with the prevailing conceptual model of deglaciation of
central British Columbia, in which mountain ranges emerge from the ice before
the plateau \citep[Fig.~7]{Fulton.1991}. If it is valid, it may suggest that
glacial lineations on the Interior Plateau of British Columbia may be of older
age than the LGM, and may have remained intact throughout the deglaciation.

@@

During deglaciation, none of the palaeo-temperature records used produces a
close agreement between the model results and the geological evidence. Although
the EPICA record yields a more realistic timing of the LGM and early
deglaciation, only the GRIP record produces a late-glacial readvance in areas
where it has been documented in the literature. Nonetheless, the
general patterns of deglaciation are consistent between both simulations driven
by the GRIP and EPICA records, and show a rapid deglaciation of the southern
half of the ice sheet, including a rapid northwards retreat across the Interior
Plateau of central British Columbia. This is followed by an opening of the ice
margin in the Liard Lowland, and a final retreat of the margin of the remaining
ice caps towards the Selwyn and, later, the Skeena mountains, which host the
last remnant of the ice sheet
during the early Holocene (10.9--9.5\,kyr).

@@

% fig:locmap
\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,kyr\,BP} (21.4\,cal\,kyr\,BP)
           \citep{Dyke.2004}. The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSE), the Selwyn and MacKenzie mountains (SMK),
           the Skeena Mountains (SM), the Coast Mountains (CM), the Rocky
           Mountains (RM) and the North Cascades (NC). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.
           \todo{Mark palaeo-ice sheets and mountain ranges on the map.
                 Mark location of the Puget Lowland.}}
  \label{fig:locmap}
\end{figure}

@@

% fig:timeseries
\begin{figure*}
  \includegraphics{timeseries}
  \caption{Temperature offset time-series from ice core and sediment core
           records (Table~\ref{tab:records}) used as palaeo-climate forcing for
           the ice sheet model (top panel), and modelled ice volume
           through the last 120\,kyr, expressed in meters of sea-level
           equivalent (bottom panel). Gray spans indicate Marine Isotope
           Stages (MIS) according to a global compilation of benthic
           \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}. Hatched
           rectangles highlight modelled ice volume extrema corresponding to
           MIS~4 (61.9--55.4\,kyr), MIS~3 (52.2--45.6\,kyr), and
           MIS~2 (last glacial maximum, 29.5--16.9\,kyr). Dotted lines
           correspond to the GRIP and EPICA 6\,km-resolution runs.}
  \label{fig:timeseries}
\end{figure*}

@@

% fig:warmbase
\begin{figure*}
  \includegraphics{warmbase}
  \caption{Modelled duration of warm-based ice cover during the last
           120\,kyr. Long ice cover durations combined with basal
           temperatures at the pressure-melting point may explain the strong
           glacial erosional imprint of the Skeena Mountains (SM) landscape.
           Hatches indicate areas that were covered by cold ice only.
           \todo{indicate location of the Skeena Mountains.}}
  \label{fig:warmbase}
\end{figure*}

@@

% fig:deglac
\begin{figure*}
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
           advance exceeded the last glacial maximum advanced are marked in
           green. Hatches denote re-advance of mountain-centred ice caps and
           and the decaying ice sheet between 14 and 10\,kyr., which is more
           pronounced in the GRIP-driven simulation.}
  \label{fig:deglac}
\end{figure*}

% fig:profiles-grip
\begin{figure}
  \includegraphics{profiles-grip}
  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
           during deglaciation (22.0--8.0\,kyr) in the GRIP 6\,km
           simulation, corresponding to the four transects indicated in
           Fig.~\ref{fig:deglac}.}
  \label{fig:profiles-grip}
\end{figure}

>>>>

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving atmospheric
  forcing, all simulations produce two major glaciation events during
  marine isotope stages~4 (61.9--55.4\,ka) and~2
  (29.5--16.9\,ka). The timing of glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional oceanic sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena Mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remains of Cordilleran ice during the
  early Holocene (10.9--9.5\,ka).

@@

Ice sheet dynamics are typically modelled using a combination of internal
deformation and basal sliding. PISM is a~shallow ice sheet model, which implies
that the balance of stresses is approximated based on their predominant
components. The Shallow Shelf Approximation (SSA) is used as a ``sliding law''
for the Shallow Ice Approximation (SIA) by adding velocity solutions of the
two approximation
\citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}\renote{
    Arjen, I don't agree with your comment. Reference to equation number(s)
    show the reader where to look. \citet{Winkelmann.etal.2011} is a rather
    long paper that also treats many other aspects of the model. Their SIA+SSA
    approach builds on \citep{Bueler.Brown.2009}, but it is slightly different.
    It is the one in use in PISM, and in my simulations.}.
Ice rheology depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
the atmospheric forcing (Sect.~\ref{sec:atm}) provides the upper boundary
condition to the ice enthalpy model. Temperature is computed subglacially to
a~depth of 3\,km\renote{
    Yes, I changed this from 1 to 3\,km as we now model longer period of times
    than in the previous paper. The penetration depth of temperature anomalies
    (i.e. skin depth) depends on the period of temperature oscillations at the
    surface. The longer the period (in our case 100\,ka), the deeper the
    temperature change. A good way to estimate what this parameter should be is
    to solve the 1-D temperature diffusion equation in the case of periodic
    surface forcing, using rock heat capacity and rock thermal conductivity
    values similar as in the numerical model. This basically forms the basis
    for my choice of the 3\,km value.},
where it is conditioned by a lower boundary geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
high spatial geothermal variability in the region
\citep{Blackwell.Richards.2004}, it is, on average, representative of available
heat flow measurements. In the low-resolution simulations, the vertical grid
consists of up to 51~enthalpy layers in the ice sheet and 31~temperature layers
in the bedrock. In the high-resolution simulations, up to 101~ice layers, and
61~bedrock layers are used.

@@

However, the exact timing of modelled MIS~2 maxima depends strongly on the
choice of applied palaeo-temperature record, which allows for a more in-depth
comparison with geological evidence for the timing of maximum Cordilleran ice
sheet extent. In the Puget lowland (Fig.~\ref{fig:locmap}), the LGM advance of
the southern Cordilleran ice sheet margin has been constrained by radiocarbon
dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
\citep{Porter.Swanson.1998}.
These dates have been confirmed by the offshore sedimentary record, which shows
an increase of glaciomarine sedimentation between 19.5 and
16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}. Radiocarbon
dating of the northern Cordilleran ice sheet margin is much less constrained
but straddles presented constraints from the southern margin. However,
cosmogenic exposure dating
places the timing of maximum CIS extent during the McConnell glaciation close
to 17\,ka \citep{Stroeven.etal.2010, Stroeven.etal.2014}.

Among the simulations presented here, only those forced with the GRIP, EPICA
and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
extents that may be compatible with these field constraints
(Fig.~\ref{fig:timeseries}, lower panel; Table~\ref{tab:extrema}).
Simulations driven by the NGRIP, ODP~1012 and ODP~1020
palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
ice sheet extents that pre-date field-based constraints by several thousands of
years. Concerning the simulations driven by oceanic records, this early
deglaciation is caused by an early warming present in the alkenone
palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
\citealp[Fig.~3]{Herbert.etal.2001}). However, this
early warming is a local effect, corresponding to a weakening of the California
current \citep{Herbert.etal.2001}. The California current, driving cold
waters southwards along the south-western coast of North America\renote{
    Not only USA but also Baja California in Mexico.},
has been shown to have weakened during each peak global glaciation (in SPECMAP)
durint the past 550\,ka, including the LGM, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
\citep{Herbert.etal.2001}.

@@

Similarly, atmospheric forcing of the model using the NGRIP palaseo-temperature
record results in an early onset of deglaciation (Fig.~\ref{fig:timeseries})
after the -22.9\,ka last glacial volume maximum (Table~\ref{tab:extrema},
incompatible with dated evidence of the last glacial maximum advance\renote{
    I actually have no explanation for that. Did not have the time to dig much
    into the ice-core literature. Any idea?}.
Hence, we focus the rest of our analysis on simulations forced by
palaeo-temperature records from GRIP and EPICA ice cores. To allow for a more
detailed comparison against the geologic record, these two simulations were
re-run using the high-resolution grid (Fig.~\ref{fig:timeseries}, lower panel,
dotted lines).

@@

The modelled ice sheet configurations corresponding to ice volume maxima during
MIS~4 are more variable (Fig.~\ref{fig:snapshots}, upper panels;
Fig.~\ref{fig:mis4}). The GRIP simulation (Fig.~\ref{fig:mis4}, left panel)
results in a modelled maximum ice sheet that closely resemble that obtained
during MIS~2, with the only major differences being a slightly less developed
northern and eastern sectors. However, the EPICA simulations produce a reduced
ice volume maximum (Fig.~\ref{fig:timeseries}), which translates in the
modelled ice sheet geometry into a significantly reduced southern sector, more
restricted ice cover on the northern and eastern flanks, and generally lower
ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).

In the GRIP simulation, some sections of the MIS~4 maximum ice margin are
slightly more extensive than their MIS~2 counterparts. This is notably the case
in the Puget Lowland, along parts of the marine margin, and parts of the north
slope of the Alaska Range. However, these differences generally correspond to a
few model grid cells. Considering the simplicity of the applied atmospheric
forcing, their interpretation would be hazardous at the scale of our
simulations.

@@

It is generally believed that the Cordilleran ice sheet formed by the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
In our simulations, major ice-dispersal centres, visible on the modelled ice
cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
Mountains (CM), the Columbia and Rocky mountains (CR), the Skeena Mountains
(SM), and the Selwyn and MacKenzie Mountains (SMK). Although the Coast,
Columbia, Rocky and Skeena Mountains (CM, CR, SM) are covered by mountain
glaciers for most of the model time, providing durable nucleation centre for
the initiation of ice sheet growth, this is not the case for the Selwyn and
MacKenzie mountains (SMK), where ice cover on the highest peaks is limited to a
fraction of the last glacial cycle. In other words, the Selwyn and MacKenzie
mountains appear as a secondary ice-dispersal centre which ``activates'' only
during the coldest periods of the last glacial cycle. The Northern Rocky
Mountains (Fig.~\ref{fig:duration}, NR), do not act as a nucleation centre,
but rather as a pinning point for the ice margin coming from the west.

Perhaps the most striking feature displayed by the distributions of modelled
ice cover is the persistence of the Skeena Mountains ice cap throughout the
entire last glacial cycle and its predominance over the other ice-dispersal
centres
(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}). Regardless of the applied
forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:snapshots},
middle panels), and serves as a nucleation centre at the onset of the glacial
readvance towards the LGM (MIS~2). The importance of residual ice for North
American glacial history leading up to the LGM has been illustrated by the
MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}\renote{
    Does this sentence really fit here? I feel that this sudden mention of the
    Laurentide ice sheet perturbates a bit the flow. Moreover, knowing that
    the Laurentide ice sheet is probably the main driver for the 100\,ka
    sea-level cycle, it would have been surprising if it had completely
    vanished at MIS~3. But for the Cordilleran ice sheet, usually thought of
    as fast-responding and intermittent, I feel that the presence of residual
    ice is less trivial.}.

@@

The modelled distribution of warm-based ice cover (Figs.~\ref{fig:warmbase}
and~\ref{fig:warmfrac}) is inevitably affected by our assumption of a constant,
70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
(Sect.~\ref{sec:overview}). However, both the Skeena Mountain area and the area
west of the MacKenzie Mountains are areas of higher-than-average geothermal
heat flux \citep{Blackwell.Richards.2004}. We can therefore expect a higher
frequency of warm-based ice cover in this areas if we were to include
spatially-variable geothermal forcing in our Cordilleran ice sheet simulations.

@@

In the North American Cordillera, the presence of lateral melt-water channels
at high elevation \citep{Margold.etal.2011, Margold.etal.2013a,
Margold.etal.2014}, and abundant esker systems at low elevation
\citep{Burke.etal.2012, Burke.etal.2012a, Perkins.etal.2013,
Margold.etal.2013}\footnote{
    Is this a good choice of references?}
indicate a rapid deglaciation. The southern and northern margins of the
Cordilleran ice sheet reached their last glacial maximum extent around
17\,ka \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998, Cosma.etal.2008,
Stroeven.etal.2010, Stroeven.etal.2014}, putting an older bound to the
onset of ice retreat. The timing of the late deglaciation is less well
constrained, but recent cosmogenic dates from north-central British Columbia
indicate that a seizable ice cap emanating from the central Coast Mountains or
the Skeena Mountains persisted into the Younger Dryas chronozone, at least
until 12.4\,ka \citep{Margold.etal.2014}.

@@

Because a general tenant in glacial geomorphology is that the majority of
landforms (lineations and eskers) are part of the deglacial envelop
\citep[terminology from][]{Kleman.etal.2006}, that is that these were formed
close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
immediately preceding deglaciation or at the time of cessation of sliding
inside the cold-base retreating margin (Fig.~\ref{fig:lastflow}). The
modelled deglacial flow patterns appear consistent between the GRIP and EPICA
simulations. They depict an active ice sheet retreat in the peripheral areas,
followed by stagnant ice decay in some of the interior regions. Several parts
of the modelling domain do not experience any basal sliding throughout the
deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
includes parts of the Interior Plateau in central British Columbia, major
portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
the MacKenzie Mountains.

Patterns of glacial lineation formed in the northern and southern sectors of
the Cordilleran ice sheet (\citealp{Prest.etal.1968};
\citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010}) show
similarities with the patterns of deglacial ice flow from numerical modelling
(Fig.~\ref{fig:lastflow}). In the northern half
of the modelling domain, modelled deglacial flow direction depict an
active downhill flow as the last remnants of the ice sheet retreat towards
elevated mountain ranges. Converging deglacial flow patterns in the Liard
Lowland, for instance (Fig.~\ref{fig:lastflow}), closely resemble that
indicated by the glacial lineations \citep[Fig.~2]{Margold.etal.2013}.

However, on the Interior Plateau of British Columbia, both
simulations predict negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), but an impressive set of glacial delineations
indicate a substantial eastward flow component of the Cordilleran ice sheet
\citep{Prest.etal.1968, Kleman.etal.2010}.
The Interior Plateau lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow patterns over the Interior
Plateau. Another explanation for the incongruent results could be that the
Interior Plateau lineation system predates deglaciation, and that deglacial
landforms are largely absent \footnote{
    \textbf{Martin, Johan}, any southwards running eskers on the plateau?}.

The
modelled deglaciation of the Interior Plateau, central British Columbia,
consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
southwards-flowing (Fig.~\ref{fig:lastflow}) non-sliding ice lobes positioned
in-between deglaciated mountain ranges
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
appears compatible with the prevailing conceptual model of deglaciation of
central British Columbia, in which mountain ranges emerge from the ice before
the plateau \citep[Fig.~7]{Fulton.1991}.

@@

During deglaciation, none of the palaeo-temperature records used produces a
close agreement between the model results and the geological evidence. Although
the EPICA record yields a more realistic timing of the LGM and early
deglaciation, only the GRIP record produces a late-glacial readvance in areas
where it has been documented in the literature. Nonetheless, the
general patterns of deglaciation are consistent between both simulations driven
by the GRIP and EPICA records, and show a rapid deglaciation of the southern
half of the ice sheet, including a rapid northwards retreat across the Interior
Plateau of central British Columbia. This is followed by an opening of the ice
margin in the Liard Lowland, and a final retreat of the margin of the remaining
ice caps towards the Selwyn and, later, the Skeena mountains, which host the
last remnant of the ice sheet
during the early Holocene (10.9--9.5\,ka).

@@

% fig:locmap
\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} (21.4\,cal\,ka)
           \citep{Dyke.2004}. The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSE), the Selwyn and MacKenzie mountains (SMK),
           the Skeena Mountains (SM), the Coast Mountains (CM), the Rocky
           Mountains (RM) and the North Cascades (NC). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.
           \todo{Mark palaeo-ice sheets and mountain ranges on the map.
                 Mark location of the Puget Lowland.}}
  \label{fig:locmap}
\end{figure}

@@

% fig:timeseries
\begin{figure*}
  \includegraphics{timeseries}
  \caption{Temperature offset time-series from ice core and sediment core
           records (Table~\ref{tab:records}) used as palaeo-climate forcing for
           the ice sheet model (top panel), and modelled ice volume
           through the last 120\,ka, expressed in meters of sea-level
           equivalent (bottom panel). Gray spans indicate Marine Isotope
           Stages (MIS) according to a global compilation of benthic
           \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}. Hatched
           rectangles highlight modelled ice volume extrema corresponding to
           MIS~4 (61.9--55.4\,ka), MIS~3 (52.2--45.6\,ka), and
           MIS~2 (last glacial maximum, 29.5--16.9\,ka). Dotted lines
           correspond to the GRIP and EPICA 6\,km-resolution runs.}
  \label{fig:timeseries}
\end{figure*}

@@

% fig:warmbase
\begin{figure*}
  \includegraphics{warmbase}
  \caption{Modelled duration of warm-based ice cover during the last
           120\,ka. Long ice cover durations combined with basal
           temperatures at the pressure-melting point may explain the strong
           glacial erosional imprint of the Skeena Mountains (SM) landscape.
           Hatches indicate areas that were covered by cold ice only.
           \todo{indicate location of the Skeena Mountains.}}
  \label{fig:warmbase}
\end{figure*}

@@

% fig:deglac
\begin{figure*}
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
           advance exceeded the last glacial maximum advanced are marked in
           green. Hatches denote re-advance of mountain-centred ice caps and
           and the decaying ice sheet between 14 and 10\,ka., which is more
           pronounced in the GRIP-driven simulation.}
  \label{fig:deglac}
\end{figure*}
